### 2024-06-15 _README.Rmd
More work

# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education

@@

- [Background](#background)  
- [The university's mission](#the-mission)  
- [What do AIs do?](#what-do-ais-do)  
- [A note on the terms of the conversation](#terms-of-debate)  
- [But really, what do AIs do?](#really-what-do-they-do)  
- [AI in teaching and learning](#ai-in-teaching-and-learning)  
- [What do GenAIs do for students?](#what-do-ais-do-for-students)  
- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)

@@

AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.

@@

Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption. This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].

@@

A mission statement is like an onion, and one needs to peel multiple layers to try to understand the core principles enunciated there. Why are things like "integrity" and "critical thought" valued? Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial]. Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022]. As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair". Is it any wonder that universities in the province are pushed to find "efficiencies"?])?

@@

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics("images/ai-bug-fix.png")
```

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023]. Not even the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html) are beyond the revolutionary all-seeing eye of AI.

```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
knitr::include_graphics("images/ai-litter-box.png")
```

The forecast that AI will improve everything (from the natural environment of the globe to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises. Examples of this abound.

@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], for one, presents three innovative use cases for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers (occasionally also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

@@

In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school). Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs). Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understanding my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). It did not take long for this prediction to turn sourer than beans in August: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.  Shortly after that, Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students". Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].]. A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):

- AI can provide personalized learning.  
- AI-powered systems can provide smart tutoring.  
- AI can automate grading.  
- AI can enable virtual classrooms.  
- AI can provide data insights for informed decisions.  
- Natural Language Processing (NLP) can enhance language learning.  
- AI can create interactive and adaptive learning materials.  
- AI can identify early signs of learning difficulties.  
- AI can aid in providing personalised professional development opportunities for teachers.
- AI can streamline administrative tasks.

Is it really possible that AIs can do all this?

@@

A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs _can_ do some things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation). Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].]. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do do?

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 70 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the importance of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many, _many_ hidden layers. As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021]. In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994]. Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing an input (e.g., a prompt) to a "statistically probable" outcome.

@@

```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 3. Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 4. Training data and model output compared"}

@@

Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data. For instance, the following figure shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response. Neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).

```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data.

```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 8. Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

There is a limit as to how flexible a model can be, and this is true of linear regression and of neural networks. Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model. Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model becomes overfit and will predict each value of the independent variable in the training dataset perfectly. The model will fail to train if there are more parameters than data points.

In a neural network trained to work with natural language, the "predictors" are the words that come before and after a certain word. Several approaches exist to _embed_ a bag of words, which means using large text corpora to calculate the frequency of co-occurrence of words. These frequencies are then converted to numerical vectors that represent the position of a word in multidimensional space. These "coordinates" in the space of the vocabulary can then be used to calculate the "distances" between words. Alas, these vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora, and the numeric values are static, even if the surrounding words in a specific prompt change. A recent breakthrough in the field of natural language processing is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after in a prompt.

@@

- AIs are neural networks, which is to say, regression models.
- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
- AIs are _very_ flexible non-linear regression models.
- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters of its newer models, but in the latest version this is estimated to be in the trillions.]. The large number of parameters makes these models _extremely_ data hungry. 
- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs. The increasingly large datasets essentially expand the space where the model could interpolate.].

@@

Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it makes sense to use this technology. To begin, we must note the number of examples of the use of AI by students. A Turnitin-sponsored  report [@tyton_partners_generative_2023] of 1,600 post secondary students and 1,000 faculty, found that 50% of students used GenAI, compared to only 22% faculty. Use of AI by students is commonly cited as a reason for universities to embrace AI [e.g., @power_3_2024]. McMaster's own internal research (based on a sample of over 1,300 students) indicates that at least 61% of students used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].

But embracing AI because the students use it is putting the cart before the horse. After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017].]. A university is supposed to be a place of education (higher, at that), and not a fast-food locale where customers get what they want.

So, what does McMaster hope to achieve with AI? From the Overview on Generative AI document:

@@

## What do GenAIs do for students?{#what-do-ais-do-for-students}

@@

```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-econometrics.png")
```

```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-cell-biology.png")
```

```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics("images/mac-library-euphemism.png")
```

@@

The preceding section illustrates how numerous tasks that AIs can notionally do in higher education are in fact poor replacements for a human-centered education--and that is even before considering the moral, ethical, and possibly legal compromises that the use of corporate-developed AIs demand (non-corporate developed AIs can be ignored: as @vandehei_behind_2024 remind us, only a handful of corporations have the resources to make AIs give the impression of competence). But what about the things that GenAIs can do _to_ students?

@@

One could ask whether there is value in doing things more slowly, less expertly at the beginning, in order to make the delicate process of thought work, so that the brain can accomplish genuinely exceptional things later, and perhaps even more efficiently.

@@

But let us assume for a moment that using AIs in their current form does becomes an in-demand tool in the short term. What is the expected longevity of skills like prompt engineering?

>>>>

# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education

@@

- [Background](#background)  
- [The university's mission](#the-mission)  
- [What do AIs do?](#what-do-ais-do)  
- [A note on the terms of the conversation](#terms-of-debate)  
- [But really, what do AIs do?](#really-what-do-they-do)  
- [AI in teaching and learning](#ai-in-teaching-and-learning)  
- [What do GenAIs do for students?](#what-can-ais-do-for-students)  
- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)

@@

AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. It seemed inevitable that universities _had to do something_ with it, at the risk of being left behind.

@@

Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption. This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn at the very top of the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023].] [@angwin_press_2024].

@@

A mission statement is like an onion, and one needs to peel multiple layers to try to understand the core principles enunciated there. Why are things like "integrity" and "critical thought" valued? Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial]. Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022], and yet, looking at the state of universities in Ontario, greater reliance on the private sector has not done much to improve the outlook. As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair". Is it any wonder that universities in the province are pushed to find "efficiencies"?])?

@@

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics("images/AI-bug-fix.png")
```

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023].

Higher-education has not been spared the onslaught of promises.

@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.] writing for Forbes, presents three innovative use cases for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers (also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

@@

In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia Business School](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school) in France, an institution that Financial Times ranks among the [top 100 in Europe](https://en.wikipedia.org/wiki/Excelia_Group#Rankings). Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs). Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been essential to understanding my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). On February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students". Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].]. A fairly comprehensive list of things that AI can do [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "potentially could", or "maybe will, at some indeterminate future"):

- AI can provide personalised learning.  
- AI-powered systems can provide smart tutoring.  
- AI can automate grading.  
- AI can enable virtual classrooms.  
- AI can provide data insights for informed decisions.  
- Natural Language Processing (NLP) can enhance language learning.  
- AI can create interactive and adaptive learning materials.  
- AI can identify early signs of learning difficulties.  
- AI can aid in providing personalised professional development opportunities for teachers.
- AI can streamline administrative tasks.

Is it possible that AIs can really do all this?

@@

A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs _can_ do some things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation). Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].]. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they do?

@@

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 65 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, all one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highligths the importance of institutions to provide the moral and ethical clarity that may elude individual instructors.].

@@

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many hidden layers. As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

@@

Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021]. In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994]. Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing a prompt to a "statistically probable" outcome.

@@

```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 2. Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 3. Training data and model output compared"}

@@

Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data. For instance, the following figure shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contigent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 4. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response; neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).

```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 5. Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 6. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

The increased complexity leads to greater flexibility, and we can see that the model makes predictions that resemble more closely the actual data.

```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 7. Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

There is a limit as to how flexible a model can be, and this is true of linear regression and of neural networks. Each additional parameter makes the model more flexible, but as a rule we cannot use more parameters than the number of data points that exist in the dataset used to train the model. Roughly, every additional parameter needs a data input to train on, and when there are as many parameters as data inputs the model is overfit and will predict each value of the independent variable in the training dataset perfectly. The model will fail to train if there are more parameters than data points.

In a neural network trained to work with natural language, the "predictors" are words that come before and after a certain word. Several approaches exist to _embed_ a bag of words, using large text corpora the co-occurrence of words is used to convert words to numerical vectors. These vectors are decontextualized, that is, they only reflect the position of a given word with respect to other words in the training corpora; however, the numeric values are static even if the surrounding words in a specific prompt change. A recent breakthrough in this field is the development of _transformers_, a network architecture architecture that updates the numeric values of a word depending on the words that come before and after.

@@

- AIs are neural networks, which is to say, regression models.
- A regression model generates outputs that are regressions to the mean (i.e., statistically probable outputs).
- AIs are _very_ flexible non-linear regression models.
- Their flexibility comes from using "deep" architectures, which means many layers with many, many parameters^[According to reports [see @griffith_gpt_4_2023], GPT-1 had 117 million parameters, GPT-2 1.5 billion, and GPT-3 175 billion parameters. OpenAI no longer reports the number of parameters, but in the latest version this is estimated to be in the trillions.].
- AIs are trained using an initial set of random values as parameters. The same training dataset can result in different models due to the randomness in the initial conditions.
- The flexibility of AIs means that outputs will vary seemingly at random with relatively small changes in the prompt. 
- Also, importantly, since regression models aim for the mean (i.e., a statistically probable output), they tend to be bad at _extrapolation_ (or thinking outside the box, if AIs could think). The further away an input (i.e., prompt) is from the space defined by the training dataset, the worse the output will tend to be^[Which is possibly why so-called hallucinations became less common as more and more data were used to train AIs. The increasingly large datasets  essentially expanded the space where the model could interpolate.].

@@

Next, I lay out some considerations about the use of AI for teaching and learning, in an attempt to find out where it might make sense to use this technology. To begin, we must note the sheer number of examples of the use of AI by students. A report sponsored by Turnitin [@tyton_partners_generative_2023] found that 50% out of 1,600 post-secondary students in the study were users of GenAI (but only 22% of 1,000 faculty). Use of AI by students is a commonly cited reason for universities to embrace AI [e.g., @power_3_2024]. McMaster's own internal research indicates that at least 61% of over 1,300 students polled, used AI during the Fall Term in 2023, 75% of them for school work [@yachouh_student_2024].

But embracing AI just because the students use it is like putting the cart before the horse. After all, smoking is banned on campus, and I do not recall the university asking the students if they smoked^["A tobacco and smoke-free campus is the next important step towards fulfilling our responsibilities as educators, healthcare professionals and to the communities we serve" said then-president Patrick Deane [@carter_mcmaster_2017].]. A university is supposed to be a place of education (higher, at that), and not a fast-food locale where customers get what they want.

What does McMaster hope to achieve with AI? From the Overview on Generative AI document:

@@

## What do GenAIs do for students?{#what-can-ais-do-for-students}

@@

```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 8. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-econometrics.png")
```

```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 9. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-cell-biology.png")
```

```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 10. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics("images/mac-library-euphemism.png")
```

@@

@@

My perspective is that it is better to do things more slowly at the beginning, to exercise the brain, so that the brain can do more exceptional things later (perhaps even more efficiently).

@@

But let us assume for a moment that AI in its current form really becomes essential. What is the expected longevity of skills like prompt engineering?
